Impact of Federal Party Leadership Changes on Donation Patterns

### 2024-12-03 paper/paper.qmd
update introduction

---
title: "Impact of Federal Party Leadership Changes on Donation Patterns"
subtitle: "An Analysis of Political Donations in Canada 2013 - 2024"
author: 
  - Maria Mangru
thanks: "Code and data are available at: [https://github.com/MariaMangru/Ontario_Political_Donors](https://github.com/MariaMangru/Ontario_Political_Donors)."
date: today
date-format: long
abstract: "This study investigates how the ruling political parties in Canada's federal and Ontario provincial governments affect the donation behaviors of individual contributors. By analyzing donation records from 2013 to 2024, the research explores whether a party being in power influences the total amount of financial support it receives. The findings reveal that opposition parties often receive more donations than those in power, highlighting donor preferences to support challengers. This insight enhances our understanding of political financing and can help parties and policymakers develop more effective fundraising strategies."
format: pdf
number-sections: true
bibliography: references.bib
editor: 
  markdown: 
    wrap: 72
---

@@

Political donations play a critical role in supporting political
campaigns and shaping electoral competition. In Canada, both federal and
provincial elections rely heavily on contributions from individuals and
organizations, which influence the political landscape and often reflect
public support for different parties. Understanding what drives donation
patterns is essential for ensuring transparency and fairness in the
democratic process and for minimizing the risk of undue influence on
policy-making.

One key question is whether a party’s position in power affects the
amount of financial support it receives. While studies in other
countries, such as the U.S.A, have shown that incumbents often benefit
from more donations, particularly from interest groups looking to
maintain influence, there has been less focus on whether this holds true
in Canada. Research like Fouirnaies and Hall’s work (REFERENCE HERE) in
the U.S. shows that donors strategically target those in power, but how
these dynamics play out in Canadian elections remains under explored.
This paper aims to fill that gap by analyzing how a party's power status
influences donor behaviour in federal and Ontario provincial elections.

Using donation data from 2013 to 2024, this study compares the financial
support received by parties in power versus those in opposition. The
analysis examines total donations by year and party while accounting for
factors like election timing and party size (major party or not). By
identifying trends in how donors respond to shifts in political power,
this research sheds light on the financial strategies behind Canada’s
democratic processes.

The results show that opposition parties tend to receive higher
donations than those in power, suggesting that donors often prioritize
supporting challengers, possibly to influence political change. These
findings align with trends observed in other political systems but
highlight unique patterns in Canada. This information is valuable not
only for political parties developing fundraising strategies but also
for policymakers and researchers seeking to ensure fairness and
accountability in political financing.

The paper is organized as follows: Section \ref{sec-data} discusses the
data sources and methodology, followed by an explanation of the
variables used in the analysis. The regression models and results are
then presented, along with interpretations. The paper concludes by
discussing the broader implications of the findings and suggesting
directions for future research.\[NEED TO UPDATE THE REFERENCES\]

# Data

@@

# Model

@@

# Results

@@

# Discussion

@@

# Appendix {#sec-appx}

>>>>

---
title: "Impact of Federal Party Leadership Changes on Donation Patterns"
subtitle: "What Data from 2013 - 2024 Reveal About Party Funding Trends"
author: 
  - Maria Mangru
thanks: "Code and data are available at: [https://github.com/MariaMangru/Ontario_Political_Donors](https://github.com/MariaMangru/Ontario_Political_Donors)."
date: today
date-format: long
abstract: "This study investigates how the ruling political parties in Canada's federal and Ontario provincial governments affect the donation behaviors of individual contributors. By analyzing donation records from 2013 to 2024, the research explores whether a party being in power influences the total amount of financial support it receives. The findings reveal that opposition parties often receive more donations than those in power, highlighting donor preferences to support challengers. This insight enhances our understanding of political financing and can help parties and policymakers develop more effective fundraising strategies."
format: pdf
number-sections: true
bibliography: references.bib
editor: 
  markdown: 
    wrap: 72
---

@@

Political donations are an important part of democratic processes, providing essential financial support for political campaigns and influencing electoral competition. In Canada, both federal and provincial elections rely on contributions from individuals and organizations. This can shape the political landscape and reflect public support for different parties. Understanding the factors that drive donation patterns is crucial for ensuring transparency and fairness in the democratic process and for minimizing the risk of undue influence on policy-making.

A key question in political finance is whether a party's position in power affects the amount of financial support it receives. Studies in other countries, such as the  @fouirnaies2014financial conducted in the United States, have shown that incumbents often benefit from more donations, particularly from interest groups seeking to maintain influence. However, there has been less focus on whether this holds true in Canada, where the political system and donation regulations differ significantly. This paper aims to fill that gap by analyzing how a party's power status influences donor behaviour in federal and Ontario provincial elections.

The primary estimand of this study is the effect of a political party's power status on the total amount of financial donations it receives from individual contributors. By comparing the financial support received by parties in power versus those in opposition, this research aims to understand whether being in power affects a party's fundraising success.

Using donation data from 2013 to 2024, this study examines total donations by year and party, accounting for factors such as election timing and party size (major party or not). The analysis employs linear regression models to quantify the relationship between power status and donation amounts, providing insights into donor motivations and the strategic considerations of political parties.

The results indicate that opposition parties tend to receive higher total donations than those in power, suggesting that donors often prioritize supporting challengers, possibly to influence political change. This information is valuable not only for political parties developing fundraising strategies but also for policymakers and researchers seeking to ensure fairness and accountability in political financing.

The paper is organized as follows: Section \ref{sec-data} discusses the data sources and methodology, including measurement and data cleaning processes. Section \ref{sec-model} presents the regression models used in the analysis, along with justification and validation. Section \ref{sec-results} outlines the results, and Section \ref{sec-discussion} discusses the implications, limitations, and suggestions for future research. Section \ref{sec-appendix} offers further detailed insights into the data, modeling approach, and methodology.

# Data {#sec-data}

@@

# Model {#sec-model}

@@

# Results {#sec-results}

@@

# Discussion {#sec-discussion}

@@

# Appendix {#sec-appendix}
